---
title: "Exploratory Analysis of Metabolic Gene Expression in AML"
subtitle: "Cross-cohort comparison of BEAT-AML and TCGA-LAML"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
editor: source
execute:
  warning: false
  message: false
---

# Introduction

This notebook presents a step-by-step exploratory analysis of metabolic gene expression in acute myeloid leukemia (AML), comparing two independent cohorts: BEAT-AML and TCGA-LAML.

The analysis focuses on genes annotated in the Human-GEM metabolic model. The main aim is to characterize metabolic expression patterns, evaluate the overlap between cohorts, and prepare the data for downstream multivariate and pathway-level analyses.

# Setup

```{r}
#| label: setup
#| include: false

library(tidyverse)

Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 100)
```

# Data loading

We load pre-processed expression matrices for BEAT-AML and TCGA-LAML, restricted to samples with matched whole-exome sequencing information.

The BEAT-AML expression matrix contains log2 TMM-CPM normalized values, while the TCGA-LAML expression matrix contains linear TMM-CPM normalized values. Both datasets use HUGO gene symbols.

Gene-level metabolic annotation from Human-GEM is used to restrict downstream analyses to genes linked to known metabolic reactions and subsystems.

```{r}
#| label: load-data
#| include: false

expr_beat_wes <- read_tsv(
  "../../data/processed/beat_wes_expression.tsv.xz",
  show_col_types = FALSE) %>%
  select(sample_id, where(~ is.numeric(.x) && sum(.x, na.rm = TRUE) > 0))

meta_beat_wes <- read_tsv(
  "../../data/processed/beat_wes_metadata.tsv",
  show_col_types = FALSE) %>%
  mutate(cohort = "BEAT", .after = 1)

expr_tcga_wes <- read_tsv(
  "../../data/processed/tcga_wes_expression.tsv.xz",
  show_col_types = FALSE) %>%
  select(sample_id, where(~ is.numeric(.x) && sum(.x, na.rm = TRUE) > 0))

meta_tcga_wes <- read_tsv(
  "../../data/processed/tcga_wes_metadata.tsv",
  show_col_types = FALSE)

gem <- read_tsv(
  "../../data/processed/human_gem_gene_subsystem_summary.tsv",
  show_col_types = FALSE)
```

# Dataset overview

Before focusing on metabolic genes, we summarize the overlap between the BEAT-AML and TCGA-LAML expression matrices. We report the number of matched RNA-WES samples, shared and cohort-specific genes, and the fraction of expression signal and zero values associated with each gene set.

```{r}
#| label: gene-and-sample-overlap-summary
#| code-fold: true
#| code-summary: "Show code"

local({
  
  inter <- intersect(colnames(expr_beat_wes)[-1], colnames(expr_tcga_wes)[-1])
  
  only_beat <- setdiff(colnames(expr_beat_wes)[-1], colnames(expr_tcga_wes)[-1])
  only_tcga <- setdiff(colnames(expr_tcga_wes)[-1], colnames(expr_beat_wes)[-1])
  
  tibble(
    cohort = c("BEAT-AML", "TCGA-LAML"),
    
    matched_rna_wes_samples = c(
      n_distinct(expr_beat_wes$sample_id),
      n_distinct(expr_tcga_wes$sample_id)
    ),
    
    total_genes = c(
      ncol(expr_beat_wes) - 1,
      ncol(expr_tcga_wes) - 1
    ),
    
    shared_genes = length(inter),
    
    cohort_specific_genes = c(
      length(only_beat),
      length(only_tcga)
    ),
    
    zero_fraction_in_shared_genes = c(
      mean(as.matrix(expr_beat_wes[, inter, drop = FALSE]) == 0, na.rm = TRUE),
      mean(as.matrix(expr_tcga_wes[, inter, drop = FALSE]) == 0, na.rm = TRUE)
    ),
    
    shared_expression_fraction = c(
      sum(as.matrix(expr_beat_wes[, inter, drop = FALSE]), na.rm = TRUE) /
        sum(as.matrix(expr_beat_wes[, -1, drop = FALSE]), na.rm = TRUE),
      
      sum(as.matrix(expr_tcga_wes[, inter, drop = FALSE]), na.rm = TRUE) /
        sum(as.matrix(expr_tcga_wes[, -1, drop = FALSE]), na.rm = TRUE)
    ),
    
    zero_fraction_in_cohort_specific_genes = c(
      if (length(only_beat) == 0) {
        NA_real_
      } else {
        mean(as.matrix(expr_beat_wes[, only_beat, drop = FALSE]) == 0, na.rm = TRUE)
      },
      
      if (length(only_tcga) == 0) {
        NA_real_
      } else {
        mean(as.matrix(expr_tcga_wes[, only_tcga, drop = FALSE]) == 0, na.rm = TRUE)
      }
    )
  ) |>
    mutate(
      zero_fraction_in_shared_genes = scales::percent(
        zero_fraction_in_shared_genes,
        accuracy = 0.01
      ),
      shared_expression_fraction = scales::percent(
        shared_expression_fraction,
        accuracy = 0.01
      ),
      zero_fraction_in_cohort_specific_genes = scales::percent(
        zero_fraction_in_cohort_specific_genes,
        accuracy = 0.01
      )
    ) |>
    knitr::kable(
      col.names = c(
        "Cohort",
        "Matched RNA-WES samples",
        "Total genes",
        "Shared genes",
        "Cohort-specific genes",
        "Zero values in shared genes (%)",
        "Total expression signal in shared genes (%)",
        "Zero values in cohort-specific genes (%)"
      ),
      align = c("l", rep("c", 7))
    )
})
```

A total of 22,657 genes are shared between BEAT-AML and TCGA-LAML. These genes account for more than 99% of the total expression signal in both cohorts and are therefore suitable for downstream comparative analyses.

BEAT-AML contains more than 10,000 additional cohort-specific genes. However, these genes contribute only about 0.8% of the total expression signal and show a high fraction of zero values, around 77%.

# Metabolic gene overview

We now focus on genes annotated as metabolic in Human-GEM. Metabolic genes are divided into shared genes, present in both BEAT-AML and TCGA-LAML, and cohort-specific genes.

```{r}
#| label: metabolic-zoom-overview
#| code-fold: true
#| code-summary: "Show code"

local({
  
  metab_shared <- Reduce(
    intersect,
    list(
      colnames(expr_beat_wes)[-1],
      colnames(expr_tcga_wes)[-1],
      gem$hugo
    )
  )
  
  metab_beat_only <- intersect(colnames(expr_beat_wes)[-1], gem$hugo) |>
    setdiff(metab_shared)
  
  metab_tcga_only <- intersect(colnames(expr_tcga_wes)[-1], gem$hugo) |>
    setdiff(metab_shared)
  
  tibble(
    cohort = c("BEAT-AML", "TCGA-LAML"),
    
    metabolic_genes = c(
      length(c(metab_shared, metab_beat_only)),
      length(c(metab_shared, metab_tcga_only))
    ),
    
    shared_metabolic_genes = length(metab_shared),
    
    cohort_specific_metabolic_genes = c(
      length(metab_beat_only),
      length(metab_tcga_only)
    ),
    
    expression_in_shared_metabolic_genes = c(
      sum(as.matrix(expr_beat_wes[, metab_shared, drop = FALSE]), na.rm = TRUE) /
        sum(as.matrix(expr_beat_wes[, -1, drop = FALSE]), na.rm = TRUE),
      
      sum(as.matrix(expr_tcga_wes[, metab_shared, drop = FALSE]), na.rm = TRUE) /
        sum(as.matrix(expr_tcga_wes[, -1, drop = FALSE]), na.rm = TRUE)
    ),
    
    expression_in_cohort_specific_metabolic_genes = c(
      if (length(metab_beat_only) == 0) {
        NA_real_
      } else {
        sum(as.matrix(expr_beat_wes[, metab_beat_only, drop = FALSE]), na.rm = TRUE) /
          sum(as.matrix(expr_beat_wes[, -1, drop = FALSE]), na.rm = TRUE)
      },
      
      if (length(metab_tcga_only) == 0) {
        NA_real_
      } else {
        sum(as.matrix(expr_tcga_wes[, metab_tcga_only, drop = FALSE]), na.rm = TRUE) /
          sum(as.matrix(expr_tcga_wes[, -1, drop = FALSE]), na.rm = TRUE)
      }
    )
  ) |>
    mutate(
      expression_in_shared_metabolic_genes = scales::percent(
        expression_in_shared_metabolic_genes,
        accuracy = 0.01
      ),
      expression_in_cohort_specific_metabolic_genes = scales::percent(
        expression_in_cohort_specific_metabolic_genes,
        accuracy = 0.01
      )
    ) |>
    knitr::kable(
      col.names = c(
        "Cohort",
        "Metabolic genes",
        "Shared metabolic genes",
        "Cohort-specific metabolic genes",
        "Total expression signal in shared metabolic genes (%)",
        "Total expression signal in cohort-specific metabolic genes (%)"
      ),
      align = c("l", rep("c", 5))
    )
})
```

Most metabolic genes are shared between BEAT-AML and TCGA-LAML. Cohort-specific metabolic genes account for only a small fraction of the total expression signal, supporting the use of shared metabolic genes for downstream comparative analyses.

# Create long-format expression table

As a starting point, we retain only genes shared between BEAT-AML and TCGA-LAML. These shared genes account for more than 99% of the total expression signal in both cohorts and provide a common feature for downstream analyses.

```{r}
#| label: create-long-aml-table
#| code-fold: true
#| code-summary: "Show code"

long_shared_aml <- local({
  
  shared_all_genes <- intersect(colnames(expr_beat_wes)[-1], colnames(expr_tcga_wes)[-1])
  
  bind_rows(
    expr_beat_wes |>
      select(sample_id, all_of(shared_all_genes)) |>
      pivot_longer(-sample_id, names_to = "hugo", values_to = "log2_tmm_cpm") |>
      mutate(tmm_cpm = 2^log2_tmm_cpm - 1, .before = "log2_tmm_cpm") |>
      left_join(gem, by = "hugo") |>
      left_join(select(meta_beat_wes, cohort, sample_id, Patient_Id, Gender, Disease_Type), by = "sample_id"),
    
    expr_tcga_wes |>
      select(sample_id, all_of(shared_all_genes)) |>
      pivot_longer(-sample_id, names_to = "hugo", values_to = "tmm_cpm") |>
      mutate(log2_tmm_cpm = log2(tmm_cpm + 1)) |>
      left_join(gem, by = "hugo") |>
      left_join(meta_tcga_wes, by = "sample_id")
  ) |>
    mutate(
      hugo_fct = if_else(hugo %in% gem$hugo, hugo, "non_metabolic"),
      subsystems = if_else(hugo_fct == "non_metabolic", "non_metabolic", subsystems),
      pathway_class = if_else(hugo_fct == "non_metabolic", "non_metabolic", pathway_class)
    ) |>
    relocate(hugo_fct, .after = "hugo")
})

glimpse(long_shared_aml)
```

We then inspect the total expression signal per sample to identify possible outliers before moving to pathway-level summaries.

```{r, fig.width=6, fig.height=2.5}
sample_depth <- long_shared_aml %>%
  group_by(cohort, sample_id) %>%
  summarise(sample_tmm_cpm = sum(tmm_cpm),
            sample_log2_tmm_cpm = sum(log2_tmm_cpm),
            .groups = "drop")

sample_depth %>%
  arrange(sample_log2_tmm_cpm)

sample_depth %>%
  arrange(desc(sample_log2_tmm_cpm))

sample_depth %>%
  ggplot(aes(x = sample_log2_tmm_cpm, fill = cohort)) +
  geom_histogram(color = "lightgray", bins = 20) +
  theme_bw() +
  facet_wrap(~cohort, ncol = 2, scales = "free")

rm(sample_depth)
```

```{r}
long_shared_aml <- long_shared_aml %>%
  filter(sample_id != "15-00300")

meta_beat_wes <- meta_beat_wes %>%
  filter(sample_id != "15-00300")

expr_beat_wes <- expr_beat_wes %>%
  filter(sample_id != "15-00300")
```

Sample `15-00300` is removed because it shows an unusually low expression signal compared with the remaining BEAT-AML samples.

```{r}
long_shared_aml %>%
  group_by(cohort) %>%
  skimr::skim(log2_tmm_cpm)
```

# Select main pathway classes

To simplify pathway-level visualization, we retain the most represented pathway classes and group the remaining metabolic pathways as `other_metabolic`.

```{r}
#| label: select-main-pathway-classes
#| code-fold: true
#| code-summary: "Show code"

top_pathway_class <- long_shared_aml |>
  filter(pathway_class != "non_metabolic") |>
  group_by(sample_id, pathway_class) |>
  summarise(abun_metab = sum(log2_tmm_cpm), .groups = "drop") |>
  group_by(sample_id) |>
  mutate(rela_metab = abun_metab / sum(abun_metab)) |>
  ungroup() |>
  summarise(total_rela = sum(rela_metab), .by = pathway_class) |>
  slice_max(total_rela, n = 20) |>
  pull(pathway_class)

long_shared_aml <- long_shared_aml |>
  mutate(
    pathway_class_fct = case_when(
      is.na(pathway_class) ~ NA_character_,
      pathway_class == "non_metabolic" ~ "non_metabolic",
      pathway_class %in% top_pathway_class ~ pathway_class,
      TRUE ~ "other_metabolic"
    ),
    pathway_class_fct = factor(
      pathway_class_fct,
      levels = c(top_pathway_class, "other_metabolic", "non_metabolic")
    )
  ) |>
  relocate(pathway_class_fct, .after = pathway_class)

rm(top_pathway_class)
```

The resulting `pathway_class_fct` variable is used only for visualization. It preserves the main metabolic pathway classes and groups less represented ones into `other_metabolic`.

```{r}
#| code-fold: true
#| code-summary: "Show plot function"

plot_stacked_bar <- function(data,
                             include_non_metabolic = TRUE,
                             relative              = FALSE,
                             title                 = NULL) {

  if (!include_non_metabolic) {
    data <- data |> filter(pathway_class_fct != "non_metabolic")
  }

  data <- data |>
    group_by(cohort, sample_id, pathway_class_fct) |>
    summarise(log2_tmm_cpm = sum(log2_tmm_cpm), .groups = "drop")

  if (relative) {
    data <- data |>
      group_by(cohort, sample_id) |>
      mutate(y_var = log2_tmm_cpm / sum(log2_tmm_cpm)) |>
      ungroup()
    y_label <- "Relative pathway signal"
  } else {
    data <- data |> mutate(y_var = log2_tmm_cpm)
    y_label <- "log2(TMM-CPM+1)"
  }

  data |>
    ggplot(aes(x = sample_id, y = y_var, fill = pathway_class_fct)) +
    geom_col() +
    theme_minimal() +
    theme(
      axis.text.x     = element_blank(),
      axis.ticks.x    = element_blank(),
      legend.position = "top"
    ) +
    labs(
      x     = "Samples",
      y     = y_label,
      fill  = "Pathway class",
      title = title
    ) +
    facet_wrap(~cohort, nrow = 2, scales = "free") +
    ggqualpal::scale_fill_qualpal()
}
```

# Pathway profile visualization

The following plots compare pathway-level profiles across samples. Absolute plots show the summed expression signal, while relative plots show the proportion of each pathway class within each sample.

::: {.panel-tabset}

## Absolute — all shared genes

```{r, fig.width=14, fig.height=8}
long_shared_aml |> 
  select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = TRUE, relative = FALSE)
```

## Absolute — metabolic only

```{r, fig.width=14, fig.height=8}
long_shared_aml |> 
  select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = FALSE, relative = FALSE)
```

## Relative — all shared genes

```{r, fig.width=14, fig.height=8}
long_shared_aml |> 
  select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(relative = TRUE)
```

## Relative — metabolic only

```{r, fig.width=14, fig.height=8}
long_shared_aml |> 
  select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = FALSE, relative = TRUE)
```

:::

The relative metabolic-only profiles are the most useful view for comparing pathway composition across samples, since they remove the dominant contribution of non-metabolic genes.

Overall, pathway profiles appear very stable across samples and across both cohorts. No major differences are evident in either the absolute pathway signal or the relative pathway composition.

```{r}
rm(plot_stacked_bar)
```


# Metabolic matrix construction and batch correction

From this point onward, we restrict the analysis to metabolic genes only and build the matrices used for dimensionality reduction and following analysis in wide format.

```{r}
#| label: metabolic-selection-and-matrices
#| code-fold: true
#| code-summary: "Show code"

long_aml_metab <- long_shared_aml |>
  filter(hugo_fct != "non_metabolic")

wide_metab <- local({
  
  wide_all <- long_aml_metab |>
    group_by(hugo) |>
    filter(n_distinct(cohort) == 2) |>
    ungroup() |>
    select(sample_id, cohort, Disease_Type, Gender, hugo_fct, log2_tmm_cpm) |>
    pivot_wider(names_from = hugo_fct, values_from = log2_tmm_cpm)
  
  meta_beat <- meta_beat_wes |>
    filter(sample_id %in% wide_all$sample_id) |>
    arrange(sample_id)
  
  meta_tcga <- meta_tcga_wes |>
    filter(sample_id %in% wide_all$sample_id) |>
    arrange(sample_id)
  
  wide_beat <- wide_all |>
    filter(cohort == "BEAT") |>
    arrange(sample_id)
  
  wide_tcga <- wide_all |>
    filter(cohort == "TCGA") |>
    arrange(sample_id)
  
  wide_all <- bind_rows(wide_beat, wide_tcga)
  
  meta_aml <- wide_all |>
    select(sample_id, cohort, Disease_Type, Gender)
  
  metab_genes <- setdiff(colnames(wide_all), colnames(meta_aml))
  
  mat_aml <- wide_all |>
    select(sample_id, all_of(metab_genes)) |>
    column_to_rownames("sample_id") |>
    as.matrix()
  
  mat_beat <- wide_beat |>
    select(sample_id, all_of(metab_genes)) |>
    column_to_rownames("sample_id") |>
    as.matrix()
  
  mat_tcga <- wide_tcga |>
    select(sample_id, all_of(metab_genes)) |>
    column_to_rownames("sample_id") |>
    as.matrix()
  
  stopifnot(identical(rownames(mat_beat), meta_beat$sample_id))
  stopifnot(identical(rownames(mat_tcga), meta_tcga$sample_id))
  stopifnot(identical(rownames(mat_aml), meta_aml$sample_id))
  
  list(
    wide_all = wide_all,
    meta_aml = meta_aml,
    meta_beat = meta_beat,
    meta_tcga = meta_tcga,
    mat_aml = mat_aml,
    mat_beat = mat_beat,
    mat_tcga = mat_tcga
  )
})

wide_metab$mat_corrected <- t(limma::removeBatchEffect(
  t(wide_metab$mat_aml),
  batch  = wide_metab$meta_aml$cohort,
  design = model.matrix(~ Disease_Type + Gender, data = wide_metab$meta_aml)
))
```

The combined matrix is used to inspect cohort-level structure before and after batch correction. Batch correction is used here only as an exploratory step to evaluate whether the dominant separation between BEAT-AML and TCGA-LAML is mainly driven by cohort-specific effects.

# Dimensionality reduction

## PCA

We first use PCA to visualize the main axes of variation in metabolic gene expression.

```{r}
#| label: pca-plot-function
#| code-fold: true
#| code-summary: "Show PCA plot function"

pca_plot <- function(pca_obj, color_vec, title, palette = NULL) {
  
  n_levels <- length(unique(color_vec))
  
  if (is.null(palette)) {
    palette <- ggpubr::get_palette("jco", n_levels)
  }
  
  factoextra::fviz_pca_biplot(
    pca_obj,
    geom.ind     = "point",
    col.ind      = color_vec,
    addEllipses  = n_levels > 1,
    ellipse.type = "norm",
    geom.var     = c("arrow", "text"),
    col.var      = "darkgray",
    select.var   = list(cos2 = 20),
    repel        = TRUE,
    palette      = palette,
    legend.title = "",
    title        = title
  ) +
    theme_bw() +
    coord_equal()
}
```

```{r}
#| label: run-pca
#| code-fold: true
#| code-summary: "Show code"

res_pca <- list(
  combined  = prcomp(wide_metab$mat_aml,       center = TRUE, scale. = FALSE),
  corrected = prcomp(wide_metab$mat_corrected, center = TRUE, scale. = FALSE),
  beat      = prcomp(wide_metab$mat_beat,      center = TRUE, scale. = FALSE),
  tcga      = prcomp(wide_metab$mat_tcga,      center = TRUE, scale. = FALSE)
)
```

::: {.panel-tabset}

## Combined

```{r}
#| label: pca-combined
#| fig-width: 10
#| fig-height: 7
#| message: false
#| code-fold: true
#| code-summary: "Show code"

pca_plot(
  res_pca$combined,
  color_vec = wide_metab$meta_aml$cohort,
  title     = "PCA — combined cohorts (no correction)",
  palette   = c("#E64B35", "#4DBBD5")
)
```

## Batch corrected

```{r}
#| label: pca-batch-corrected
#| fig-width: 10
#| fig-height: 7
#| message: false
#| code-fold: true
#| code-summary: "Show code"

pca_plot(
  res_pca$corrected,
  color_vec = wide_metab$meta_aml$cohort,
  title     = "PCA — combined cohorts (batch corrected)",
  palette   = c("#E64B35", "#4DBBD5")
)
```

## BEAT-AML

```{r}
#| label: pca-beat
#| fig-width: 10
#| fig-height: 7
#| message: false
#| code-fold: true
#| code-summary: "Show code"

pca_plot(
  res_pca$beat,
  color_vec = wide_metab$meta_beat$Disease_Type,
  title     = "PCA — BEAT-AML"
)
```

## TCGA-LAML

```{r}
#| label: pca-tcga
#| fig-width: 10
#| fig-height: 7
#| message: false
#| code-fold: true
#| code-summary: "Show code"

pca_plot(
  res_pca$tcga,
  color_vec = wide_metab$meta_tcga$Disease_Type,
  title     = "PCA — TCGA-LAML"
)
```

:::

The uncorrected combined PCA is used to assess whether the two cohorts separate before correction. The batch-corrected PCA provides a comparison after removing cohort-associated effects. The cohort-specific PCA plots are then used to inspect the internal structure of BEAT-AML and TCGA-LAML separately.

```{r}
rm(pca_plot)
```


## UMAP

```{r}
library(uwot)

umap_plot <- function(umap_mat, color_vec, title, palette = NULL) {
  
  n_levels <- length(unique(color_vec))
  
  if (is.null(palette)) {
    palette <- ggpubr::get_palette("jco", n_levels)
  }
  
  p <- tibble(
    UMAP1 = umap_mat[, 1],
    UMAP2 = umap_mat[, 2],
    group = color_vec
  ) %>%
    ggplot(aes(UMAP1, UMAP2, color = group)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = palette) +
    theme_bw() +
    coord_equal() +
    labs(title = title, color = "")
  
  if (n_levels > 1) {
    p <- p + stat_ellipse(type = "norm", linewidth = 0.8)
  }
  
  p
}
```

```{r}
set.seed(42)
res_umap <- list(
  combined  = uwot::umap(wide_metab$mat_aml, n_neighbors = 15, min_dist = 0.1),
  corrected = uwot::umap(wide_metab$mat_corrected, n_neighbors = 15, min_dist = 0.1),
  beat      = uwot::umap(wide_metab$mat_beat, n_neighbors = 15, min_dist = 0.1),
  tcga      = uwot::umap(wide_metab$mat_tcga,n_neighbors = 15, min_dist = 0.1))
```

::: {.panel-tabset}

## Combined

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$combined,
          color_vec = wide_metab$meta_aml$cohort,
          title     = "UMAP — combined cohorts (no correction)",
          palette   = c("#E64B35", "#4DBBD5"))
```

## Batch corrected

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$corrected,
          color_vec = wide_metab$meta_aml$cohort,
          title     = "UMAP — combined cohorts (batch corrected)",
          palette   = c("#E64B35", "#4DBBD5"))
```

## BEAT-AML

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$beat,
          color_vec = meta_beat_wes$Disease_Type,
          title     = "UMAP — BEAT-AML")
```

## TCGA-AML

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$tcga,
          color_vec = meta_tcga_wes$Disease_Type,
          title     = "UMAP — TCGA-LAML")
```

:::

```{r}
rm(umap_plot)
```


# Heatmaps

```{r}
pheatmap_params <- list(
  # Annotation bars
  anno_beat = wide_metab$meta_beat %>%
    select(sample_id, Disease_Type, Gender) %>%
    column_to_rownames("sample_id"),
  
  anno_tcga = wide_metab$meta_tcga %>%
    select(sample_id, Disease_Type, Gender) %>%
    column_to_rownames("sample_id"),
  
  # Common heatmap arguments
  heatmap_args = list(
    scale        = "column",        # z-score per gene
    cluster_rows = TRUE,            # cluster samples
    cluster_cols = TRUE,            # cluster genes
    show_rownames = FALSE,          # too many samples to label
    show_colnames = FALSE,
    color        = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
    border_color = NA,
    cellwidth     = .5, 
    cellheight     = 2
  )
)
```


::: {.panel-tabset}

## BEAT

```{r, fig.width=24, fig.height=12}
do.call(pheatmap::pheatmap, c(
  list(mat            = wide_metab$mat_beat,
       main           = "BEAT",
       annotation_row = pheatmap_params$anno_beat),
  pheatmap_params$heatmap_args))
```

## TCGA

```{r, fig.width=24, fig.height=12}
do.call(pheatmap::pheatmap, c(
  list(mat            = wide_metab$mat_tcga,
       main           = "TCGA",
       annotation_row = pheatmap_params$anno_tcga),
  pheatmap_params$heatmap_args))
```

:::

## Differential Expression Analysis

We test which HVGs are differentially expressed across Disease Types using `limma`,
which applies empirical Bayes moderation to stabilize variance estimates across
genes.

We test which HVGs are differentially expressed across Disease Types using `limma`,
which applies empirical Bayes moderation to stabilize variance estimates across
genes — particularly important with the sample sizes available here.

### BEAT-AML

```{r}
# Design matrix — one column per group, first level absorbed into intercept
meta_beat_wes <- meta_beat_wes %>%
  mutate(Disease_Type_fct = fct_collapse(Disease_Type,
    therapy_related = c("t-AML", "t-MNs")))

design_beat <- model.matrix(~ Disease_Type_fct, data = meta_beat)
fit    <- lmFit(t(mat_beat[, hvg_genes_any]), design_beat)
fit    <- eBayes(fit)
```

```{r}
results_beat <- topTable(fit, sort.by = "F", n = Inf) %>%
  rownames_to_column("hugo") %>%
  as_tibble()
```

```{r}
# Significant genes at FDR < 0.05
sig_beat <- results_beat %>% filter(adj.P.Val < 0.01)

glue::glue("{nrow(sig_beat)} significant HVGs at FDR < 0.01")
```

```{r}
# Bar plot — top genes by F-statistic
results_beat %>%
  filter(adj.P.Val < 0.01) %>%
  ggplot(aes(x = reorder(hugo, F), y = F)) +
  geom_col(fill = "#4DBBD5") +
  coord_flip() +
  theme_bw() +
  labs(x     = NULL,
       y     = "F-statistic",
       title = "Significant HVGs by Disease Type — BEAT-AML")
```

```{r, fig.width=12, fig.height=9}
long_aml_metab %>%
    filter(cohort == "BEAT", hugo %in% sig_beat$hugo) %>%
    mutate(Disease_Type = fct_collapse(Disease_Type,
                                       therapy_related = c("t-AML", "t-MNs")),
           hugo = factor(hugo, levels = sig_beat$hugo)) %>%
    ggplot(aes(x = Disease_Type, y = log2_tmm_cpm, fill = Disease_Type)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.size = 0.5) +
    facet_wrap(~ hugo, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = NULL, y = "log2(TMM-CPM)")
```

## WGCNA

```{r}
wes_beat <- read_tsv("../../data/processed/wes_wide.tsv",
                     show_col_types = FALSE) %>%
  filter(sample_id %in% meta_beat$sample_id)   # keep only BEAT-AML samples

# Traits matrix — Disease_Type + Gender + WES mutations
traits <- meta_beat %>%
  select(sample_id, Disease_Type, Gender) %>%
  mutate(
    is_sec             = as.integer(Disease_Type %in% "sec"),
    is_therapy_related = as.integer(Disease_Type %in% c("t-AML", "t-MNs")),
    is_female          = as.integer(Gender == "FEMALE")
  ) %>%
  select(sample_id, is_sec, is_therapy_related, is_female) %>%
  inner_join(wes_beat, by = "sample_id") %>%
  mutate(across(where(is.logical), as.integer)) %>%  # TRUE/FALSE -> 1/0
  column_to_rownames("sample_id")

# Expression matrix aligned to traits
mat_wgcna <- mat_beat[rownames(traits), gem$hugo]

# Sanity check
identical(rownames(mat_wgcna), rownames(traits))
glue::glue("Samples: {nrow(mat_wgcna)} | Genes: {ncol(mat_wgcna)} | Traits: {ncol(traits)}")
```






























<!-- ## Highly Variable Genes -->

<!-- ### Selection -->

<!-- To focus on biologically informative signal, we restrict downstream analyses to -->
<!-- genes with meaningful expression variability across samples. We compute per-gene -->
<!-- variance within each cohort independently on log2(TMM-CPM) values. -->

<!-- ```{r} -->
<!-- tbl_var <- long_aml_metab %>% -->
<!--     group_by(cohort, hugo) %>% -->
<!--     summarise(variance = var(log2_tmm_cpm), .groups = "drop") -->

<!-- tbl_var <- long_aml_metab %>% -->
<!--     group_by(cohort, hugo) %>% -->
<!--     summarise(variance = var(log2_tmm_cpm), .groups = "drop") -->

<!-- tbl_var %>% -->
<!--     ggplot(aes(x = variance, fill = cohort, group = cohort)) + -->
<!--     geom_density(alpha = .25) + theme_bw() -->

<!-- library(effsize) -->

<!-- cliff.delta( -->
<!--   tbl_var %>% filter(cohort == "BEAT") %>% pull(variance), -->
<!--   tbl_var %>% filter(cohort == "TCGA") %>% pull(variance)) -->
<!-- ``` -->

<!-- The distributions are nearly identical between cohorts (Cliff's δ = -0.007, -->
<!-- 95% CI: -0.038 to 0.024), justifying a common threshold. We retain genes with -->
<!-- variance > 1.0 in log2 scale, corresponding to a typical sample-to-sample -->
<!-- variability of at least 2-fold in linear CPM units in at least one cohort. -->
<!-- All selected HVGs are present in both datasets. -->

<!-- ```{r} -->
<!-- hvg_genes_any <- tbl_var %>% -->
<!--   group_by(hugo) %>% -->
<!--   filter(any(variance > 1.0)) %>%  -->
<!--   ungroup() %>% pull(hugo) %>% unique() -->

<!-- hvg_genes_all <- tbl_var %>% -->
<!--   group_by(hugo) %>% -->
<!--   filter(all(variance > 1.0)) %>%  -->
<!--   ungroup() %>% pull(hugo) %>% unique() -->

<!-- all(hvg_genes_any %in% colnames(mat_tcga)) -->

<!-- tbl_var %>%  -->
<!--   filter(hugo %in% hvg_genes_any) %>%  -->
<!--   pivot_wider(names_from = "cohort", values_from = "variance") %>%  -->
<!--   mutate(diff = abs(BEAT - TCGA)) %>%  -->
<!--   filter(BEAT < 1 | TCGA < 1, diff > 2) -->

<!-- tbl_var %>% -->
<!--   filter(hugo %in% hvg_genes_any) %>% -->
<!--   pivot_wider(names_from = "cohort", values_from = "variance") %>% -->
<!--   arrange(desc(BEAT + TCGA)) -->
<!-- ``` -->

<!-- Two genes show large inter-cohort variance discrepancies: CRISP3 (var: 6.36 in -->
<!-- BEAT, 0.93 in TCGA) and HSD11B1 (var: 0.12 in BEAT, 5.37 in TCGA). Both are -->
<!-- retained under the `any` threshold and excluded under `all`. -->