---
title: "explorations 2"
format: html
editor: source
---

## Setup

```{r, message=FALSE}
library(tidyverse)
library(skimr)
library(knitr)
library(factoextra)
library(limma)
```

## Data

We load pre-processed expression matrices for BEAT-AML (217 samples, 22,000+ genes,
log2 TMM-CPM normalized, Hugo symbols) and TCGA-LAML (140 samples, 30,000+ genes,
linear TMM-CPM normalized, Hugo symbols), restricted to samples with matched WES information.
Gene-level metabolic annotation from Human-GEM restricts downstream analyses to
~2,200 metabolic genes linked to known reactions and subsystems.

```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 100)
expr_beat <- read_tsv("../../data/processed/beat_aml_expression.tsv.xz",
                     show_col_types = FALSE) %>%
  select(sample_id, where(~ is.numeric(.x) && sum(.x) > 0))

meta_beat <- read_tsv("../../data/processed/beat_aml_metadata.tsv",
                     show_col_types = FALSE) %>%
  mutate(cohort = "BEAT", .after = 1)

expr_tcga <- read_tsv("../../data/processed/tcga_aml_expression.tsv.xz",
                     show_col_types = FALSE) %>%
  select(sample_id, where(~ is.numeric(.x) && sum(.x) > 0))

meta_tcga <- read_tsv("../../data/processed/tcga_aml_metadata.tsv",
                     show_col_types = FALSE)

gem <- read_tsv("../../data/processed/human_gem_gene_subsystem_summary.tsv",
                show_col_types = FALSE)
```

```{r}
tribble(
  ~dataset,    ~sample,          ~genes,          ~sparsity,
  "expr_beat", nrow(expr_beat),  ncol(expr_beat), mean(expr_beat[,-1] == 0),
  "expr_tcga", nrow(expr_tcga),  ncol(expr_tcga), mean(expr_tcga[,-1] == 0),
  "gem",       NA,               nrow(gem),        NA
) |>
  mutate(sparsity = scales::percent(sparsity, accuracy = 0.1)) |>
  knitr::kable() |>
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "bordered"),
    full_width        = FALSE,
    font_size         = 16)
```

```{r}
# Gene sets by metabolic annotation and cohort overlap
metab_genes_beat <- intersect(colnames(expr_beat)[-1], gem$hugo)
metab_genes_tcga <- intersect(colnames(expr_tcga)[-1], gem$hugo)
shared_metab     <- intersect(metab_genes_beat, metab_genes_tcga)
only_beat        <- setdiff(metab_genes_beat, metab_genes_tcga)
only_tcga        <- setdiff(metab_genes_tcga, metab_genes_beat)
non_metab_beat   <- setdiff(colnames(expr_beat)[-1], gem$hugo)
non_metab_tcga   <- setdiff(colnames(expr_tcga)[-1], gem$hugo)

# Total abundance per cohort for relative contribution
total_beat <- sum(expr_beat[-1])
total_tcga <- sum(expr_tcga[-1])

tribble(
  ~category,             ~n_genes_beat,          ~n_genes_tcga,          ~abun_pct_beat,                                    ~abun_pct_tcga,
  "Shared metabolic",    length(shared_metab),   length(shared_metab),   sum(expr_beat[shared_metab]) / total_beat * 100,   sum(expr_tcga[shared_metab]) / total_tcga * 100,
  "Only BEAT metabolic", length(only_beat),      NA,                     sum(expr_beat[only_beat])    / total_beat * 100,   NA,
  "Only TCGA metabolic", NA,                     length(only_tcga),      NA,                                                sum(expr_tcga[only_tcga])    / total_tcga * 100,
  "Non-metabolic",       length(non_metab_beat), length(non_metab_tcga), sum(expr_beat[non_metab_beat])/ total_beat * 100,  sum(expr_tcga[non_metab_tcga])/ total_tcga * 100
) |>
  mutate(across(starts_with("abun"), ~ round(.x, 2))) |>
  knitr::kable(
    col.names = c(
      "Category",
      "Genes BEAT", "Genes TCGA",
      "% abundance BEAT", "% abundance TCGA"
    ),
    caption = "2,600 metabolic genes are shared between cohorts and used for downstream
    analysis. 224 genes are metabolic but exclusive to BEAT-AML, contributing only
    ~0.02% of total abundance - are these negligible? No metabolic genes are
    exclusive to TCGA. Non-metabolic genes dominate both cohorts by abundance
    (84% in BEAT-AML, 76% in TCGA-LAML), which is expected as Human-GEM covers
    only a subset of the transcriptome."
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "bordered"),
    full_width        = FALSE,
    font_size         = 16)
```

```{r, fig.width=12, fig.height=2.5}
expr_only_beat <- expr_beat[, c("sample_id",
                                setdiff(colnames(expr_beat), colnames(expr_tcga)))] 
expr_shared_beat <- expr_beat[, c("sample_id",
                                intersect(colnames(expr_beat), colnames(expr_tcga)))] 

ggpubr::ggarrange(
  
(colSums(expr_only_beat[, -1] > 0) / nrow(expr_only_beat)) %>%
  enframe(name = "hugo", value = "prevalence") %>%
  ggplot(aes(x = prevalence)) +
  geom_histogram(fill = "lightblue", color = "darkblue", bins = 30) + 
  theme_bw() + ggtitle("exclusive genes in beat"),

(colSums(expr_shared_beat[, -1] > 0) / nrow(expr_shared_beat)) %>%
  enframe(name = "hugo", value = "prevalence") %>%
  ggplot(aes(x = prevalence)) +
  geom_histogram(fill = "lightblue", color = "darkblue", bins = 30) + 
  theme_bw() + ggtitle("shared genes in beat"))
```

```{r}
rm(metab_genes_beat, metab_genes_tcga, shared_metab,
   only_beat, only_tcga, non_metab_beat, non_metab_tcga,
   total_beat, total_tcga, expr_only_beat, expr_shared_beat)
```

## Create Long Format tibble

```{r}
long_aml <- bind_rows(
  # beat
  expr_beat %>%
    pivot_longer(-sample_id, names_to = "hugo", values_to = "log2_tmm_cpm") %>%
    mutate(tmm_cpm = 2^log2_tmm_cpm - 1, .before = "log2_tmm_cpm") %>%
    ungroup() %>%
    left_join(gem, by = "hugo") %>%
    left_join(
      select(meta_beat, cohort, sample_id, Patient_Id, Gender, Disease_Type),
      by = "sample_id"),
  
  # tcga
  expr_tcga %>%
    pivot_longer(-sample_id, names_to = "hugo", values_to = "tmm_cpm") %>%
    mutate(log2_tmm_cpm = log2(tmm_cpm + 1)) %>%
    left_join(gem, by = "hugo") %>%
    left_join(meta_tcga, by = "sample_id")) %>%
  
  # more
  mutate(
    hugo_fct = if_else(hugo %in% gem$hugo, hugo, "non_metabolic"),
    subsystems     = if_else(hugo_fct == "non_metabolic", "non_metabolic", subsystems),
    pathway_class = if_else(hugo_fct == "non_metabolic", "non_metabolic", pathway_class)) %>%
  relocate(hugo_fct, .after = "hugo") 

glimpse(long_aml)
```

```{r, fig.width=6, fig.height=2.5}
sample_depth <- long_aml %>%
  group_by(cohort, sample_id) %>%
  summarise(sample_tmm_cpm = sum(tmm_cpm),
            sample_log2_tmm_cpm = sum(log2_tmm_cpm),
            .groups = "drop")

sample_depth %>%
  arrange(sample_log2_tmm_cpm)

sample_depth %>%
  arrange(desc(sample_log2_tmm_cpm))

sample_depth %>%
  ggplot(aes(x = sample_log2_tmm_cpm, fill = cohort)) +
  geom_histogram(color = "lightgray", bins = 30) +
  theme_bw() +
  facet_wrap(~cohort, ncol = 2)
```

```{r}
long_aml <- long_aml %>%
  filter(sample_id != "15-00300")
```

```{r}
long_aml %>%
  group_by(cohort) %>%
  skim(log2_tmm_cpm)
```

## Pathway Profiles

```{r}
top_pathway_class <- long_aml |>
  select(sample_id, pathway_class, log2_tmm_cpm) |>
  filter(pathway_class != "non_metabolic") |>
  group_by(sample_id, pathway_class) |>
  summarise(abun_metab = sum(log2_tmm_cpm), .groups = "drop") |>
  group_by(sample_id) |>
  mutate(rela_metab = abun_metab / sum(abun_metab)) |>
  ungroup() |>
  group_by(pathway_class) |>
  summarise(total_rela = sum(rela_metab), .groups = "drop") |>
  arrange(desc(total_rela)) %>%
  slice_max(total_rela, n = 20) %>%
  pull(pathway_class) 
pathway_order <- c(top_pathway_class, "other_metabolic", "non_metabolic")

long_aml <- long_aml %>%
  mutate(pathway_class_fct = case_when(
    is.na(pathway_class)                 ~ NA_character_,
    pathway_class == "non_metabolic"     ~ "non_metabolic",
    pathway_class %in% top_pathway_class ~ pathway_class,
    TRUE                                 ~ "other_metabolic")) %>%
  mutate(pathway_class_fct = factor(pathway_class_fct, levels = pathway_order)) %>%
  relocate(.after = "pathway_class")

rm(top_pathway_class, pathway_order)
```

```{r}
#| code-fold: true
#| code-summary: "Show plot function"

plot_stacked_bar <- function(data,
                             include_non_metabolic = TRUE,
                             relative              = FALSE,
                             title                 = NULL) {

  if (!include_non_metabolic) {
    data <- data |> filter(pathway_class_fct != "non_metabolic")
  }

  data <- data |>
    group_by(cohort, sample_id, pathway_class_fct) |>
    summarise(log2_tmm_cpm = sum(log2_tmm_cpm), .groups = "drop")

  if (relative) {
    data <- data |>
      group_by(cohort, sample_id) |>
      mutate(y_var = log2_tmm_cpm / sum(log2_tmm_cpm)) |>
      ungroup()
    y_label <- "Relative abundance"
  } else {
    data <- data |> mutate(y_var = log2_tmm_cpm)
    y_label <- "log2(TMM-CPM+1)"
  }

  data |>
    ggplot(aes(x = sample_id, y = y_var, fill = pathway_class_fct)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x     = element_blank(),
          axis.ticks.x    = element_blank(),
          legend.position = "top") +
    labs(x     = "Samples",
         y     = y_label,
         fill  = "Pathway class",
         title = title) +
    facet_wrap(~cohort, nrow = 2, scales = "free") +
    ggqualpal::scale_fill_qualpal()
}
```

::: {.panel-tabset}

## Absolute — all genes

```{r, fig.width=14, fig.height=8}
long_aml |> select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = TRUE, relative = FALSE)
```

## Absolute — metabolic only

```{r, fig.width=14, fig.height=8}
long_aml |> select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = FALSE, relative = FALSE)
```

## Relative — all genes

```{r, fig.width=14, fig.height=8}
long_aml |> select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(relative = TRUE)
```

## Relative — metabolic only

```{r, fig.width=14, fig.height=8}
long_aml |> select(cohort, sample_id, pathway_class_fct, log2_tmm_cpm) |>
  plot_stacked_bar(include_non_metabolic = FALSE, relative = TRUE)
```

:::

## Metabolic Selection and batch effects

i will consider only metabolic pathways from now on

```{r}
long_aml_metab <- long_aml %>%
  filter(hugo_fct != "non_metabolic")
```

```{r}
wide_aml_metab <- long_aml_metab %>%
  group_by(hugo) %>%
  filter(n_distinct(cohort) == 2) %>%     
  ungroup() %>%
  select(sample_id, cohort, hugo_fct, Disease_Type, Gender, log2_tmm_cpm) %>%
  pivot_wider(names_from = hugo_fct, values_from = log2_tmm_cpm)

mat_aml <- wide_aml_metab %>%
  select(sample_id, DPM1:last_col()) %>%
  column_to_rownames("sample_id") %>% as.matrix()

mat_beat <- wide_aml_metab %>%
  filter(cohort == "BEAT") %>%
  select(sample_id, DPM1:last_col()) %>%
  column_to_rownames("sample_id") %>% as.matrix()

mat_tcga <- wide_aml_metab %>%
  filter(cohort == "TCGA") %>%
  select(sample_id, DPM1:last_col()) %>%
  column_to_rownames("sample_id") %>% as.matrix()

meta_aml <- wide_aml_metab %>%
  select(sample_id:Gender)

mat_corrected <- t(limma::removeBatchEffect(
  t(mat_aml),
  batch  = meta_aml$cohort,
  design = model.matrix(~ Disease_Type + Gender, data = meta_aml)))
```

## Dimensionality Reduction

### PCA

```{r}
pca_plot <- function(pca_obj, color_vec, title, palette = NULL) {
  
  n_levels <- length(unique(color_vec))
  
  if (is.null(palette)) {
    palette <- ggpubr::get_palette("jco", n_levels)
  }
  
  fviz_pca_biplot(
    pca_obj,
    geom.ind     = "point",
    col.ind      = color_vec,
    addEllipses  = n_levels > 1,   # no ellipse with a single group
    ellipse.type = "norm",
    geom.var     = c("arrow", "text"),
    col.var      = "darkgray",
    select.var   = list(cos2 = 20),
    repel        = TRUE,
    palette      = palette,
    legend.title = "",
    title        = title
  ) + theme_bw() + coord_equal()
}
```

```{r}
res_pca <- list(
  combined  = prcomp(mat_aml,       center = TRUE, scale. = FALSE),
  corrected = prcomp(mat_corrected, center = TRUE, scale. = FALSE),
  beat      = prcomp(mat_beat,      center = TRUE, scale. = FALSE),
  tcga      = prcomp(mat_tcga,      center = TRUE, scale. = FALSE))
```

::: {.panel-tabset}

## Combined

```{r, fig.width=10, fig.height=7, message=FALSE}
pca_plot(res_pca$combined,
         color_vec = meta_aml$cohort,
         title     = "PCA — combined cohorts (no correction)",
         palette   = c("#E64B35", "#4DBBD5"))
```

## Batch corrected

```{r, fig.width=10, fig.height=7, message=FALSE}
pca_plot(res_pca$corrected,
         color_vec = meta_aml$cohort,
         title     = "PCA — combined cohorts (batch corrected)",
         palette   = c("#E64B35", "#4DBBD5"))
```

## BEAT

```{r, fig.width=10, fig.height=7, message=FALSE}
pca_plot(res_pca$beat,
         color_vec = meta_beat$Disease_Type,
         title     = "PCA — BEAT-AML")
```

## TCGA

```{r, fig.width=10, fig.height=7, message=FALSE}
pca_plot(res_pca$tcga,
         color_vec = meta_tcga$Disease_Type,
         title     = "PCA — TCGA-LAML")
```

:::

### UMAP

### UMAP

```{r}
library(uwot)

umap_plot <- function(umap_mat, color_vec, title, palette = NULL) {
  
  n_levels <- length(unique(color_vec))
  
  if (is.null(palette)) {
    palette <- ggpubr::get_palette("jco", n_levels)
  }
  
  p <- tibble(
    UMAP1 = umap_mat[, 1],
    UMAP2 = umap_mat[, 2],
    group = color_vec
  ) %>%
    ggplot(aes(UMAP1, UMAP2, color = group)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = palette) +
    theme_bw() +
    coord_equal() +
    labs(title = title, color = "")
  
  if (n_levels > 1) {
    p <- p + stat_ellipse(type = "norm", linewidth = 0.8)
  }
  
  p
}
```

```{r}
set.seed(42)
res_umap <- list(
  combined  = uwot::umap(mat_aml, n_neighbors = 15, min_dist = 0.1),
  corrected = uwot::umap(mat_corrected, n_neighbors = 15, min_dist = 0.1),
  beat      = uwot::umap(mat_beat, n_neighbors = 15, min_dist = 0.1),
  tcga      = uwot::umap(mat_tcga,n_neighbors = 15, min_dist = 0.1))
```

::: {.panel-tabset}

## Combined

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$combined,
          color_vec = meta_aml$cohort,
          title     = "UMAP — combined cohorts (no correction)",
          palette   = c("#E64B35", "#4DBBD5"))
```

## Batch corrected

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$corrected,
          color_vec = meta_aml$cohort,
          title     = "UMAP — combined cohorts (batch corrected)",
          palette   = c("#E64B35", "#4DBBD5"))
```

## BEAT-AML

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$beat,
          color_vec = meta_beat$Disease_Type,
          title     = "UMAP — BEAT-AML")
```

## TCGA-AML

```{r, fig.width=10, fig.height=7}
umap_plot(res_umap$tcga,
          color_vec = meta_tcga$Disease_Type,
          title     = "UMAP — TCGA-LAML")
```

:::

## Highly Variable Genes

### Selection

To focus on biologically informative signal, we restrict downstream analyses to
genes with meaningful expression variability across samples. We compute per-gene
variance within each cohort independently on log2(TMM-CPM) values.

```{r}
tbl_var <- long_aml_metab %>%
    group_by(cohort, hugo) %>%
    summarise(variance = var(log2_tmm_cpm), .groups = "drop")

tbl_var <- long_aml_metab %>%
    group_by(cohort, hugo) %>%
    summarise(variance = var(log2_tmm_cpm), .groups = "drop")

tbl_var %>%
    ggplot(aes(x = variance, fill = cohort, group = cohort)) +
    geom_density(alpha = .25) + theme_bw()

library(effsize)

cliff.delta(
  tbl_var %>% filter(cohort == "BEAT") %>% pull(variance),
  tbl_var %>% filter(cohort == "TCGA") %>% pull(variance))
```

The distributions are nearly identical between cohorts (Cliff's δ = -0.007,
95% CI: -0.038 to 0.024), justifying a common threshold. We retain genes with
variance > 1.0 in log2 scale, corresponding to a typical sample-to-sample
variability of at least 2-fold in linear CPM units in at least one cohort.
All selected HVGs are present in both datasets.

```{r}
hvg_genes_any <- tbl_var %>%
  group_by(hugo) %>%
  filter(any(variance > 1.0)) %>% 
  ungroup() %>% pull(hugo) %>% unique()

hvg_genes_all <- tbl_var %>%
  group_by(hugo) %>%
  filter(all(variance > 1.0)) %>% 
  ungroup() %>% pull(hugo) %>% unique()

all(hvg_genes_any %in% colnames(mat_tcga))

tbl_var %>% 
  filter(hugo %in% hvg_genes_any) %>% 
  pivot_wider(names_from = "cohort", values_from = "variance") %>% 
  mutate(diff = abs(BEAT - TCGA)) %>% 
  filter(BEAT < 1 | TCGA < 1, diff > 2)

tbl_var %>%
  filter(hugo %in% hvg_genes_any) %>%
  pivot_wider(names_from = "cohort", values_from = "variance") %>%
  arrange(desc(BEAT + TCGA))
```

Two genes show large inter-cohort variance discrepancies: CRISP3 (var: 6.36 in
BEAT, 0.93 in TCGA) and HSD11B1 (var: 0.12 in BEAT, 5.37 in TCGA). Both are
retained under the `any` threshold and excluded under `all`.

### Heatmaps

```{r}
library(pheatmap)

beat_idx <- which(meta_aml$cohort == "BEAT")
tcga_idx <- which(meta_aml$cohort == "TCGA")

# Annotation bars
anno_beat <- meta_aml %>%
  filter(cohort == "BEAT") %>%
  select(sample_id, Disease_Type, Gender) %>%
  column_to_rownames("sample_id")

anno_tcga <- meta_aml %>%
  filter(cohort == "TCGA") %>%
  select(sample_id, Disease_Type, Gender) %>%
  column_to_rownames("sample_id")

# Common heatmap arguments
heatmap_args <- list(
  scale        = "column",        # z-score per gene
  cluster_rows = TRUE,            # cluster samples
  cluster_cols = TRUE,            # cluster genes
  show_rownames = FALSE,          # too many samples to label
  show_colnames = FALSE,
  color        = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  border_color = NA,
  cellwidth     = 2, 
  cellheight     = 2
)
```


::: {.panel-tabset}

## BEAT-AML

```{r, fig.width=16, fig.height=12}
do.call(pheatmap, c(
  list(mat            = mat_aml[beat_idx, hvg_genes_any],
       main           = "HVG expression — BEAT-AML",
       annotation_row = anno_beat),
  heatmap_args))
```

## TCGA-LAML

```{r, fig.width=16, fig.height=12}
do.call(pheatmap, c(
  list(mat            = mat_aml[tcga_idx, hvg_genes_any],
       main           = "HVG expression — TCGA-LAML",
       annotation_row = anno_tcga),
  heatmap_args))
```

:::

## Differential Expression Analysis

We test which HVGs are differentially expressed across Disease Types using `limma`,
which applies empirical Bayes moderation to stabilize variance estimates across
genes.

We test which HVGs are differentially expressed across Disease Types using `limma`,
which applies empirical Bayes moderation to stabilize variance estimates across
genes — particularly important with the sample sizes available here.

### BEAT-AML

```{r}
# Design matrix — one column per group, first level absorbed into intercept
meta_beat <- meta_beat %>%
  mutate(Disease_Type_fct = fct_collapse(Disease_Type,
    therapy_related = c("t-AML", "t-MNs")))

design_beat <- model.matrix(~ Disease_Type_fct, data = meta_beat)
fit    <- lmFit(t(mat_beat[, hvg_genes_any]), design_beat)
fit    <- eBayes(fit)
```

```{r}
results_beat <- topTable(fit, sort.by = "F", n = Inf) %>%
  rownames_to_column("hugo") %>%
  as_tibble()
```

```{r}
# Significant genes at FDR < 0.05
sig_beat <- results_beat %>% filter(adj.P.Val < 0.01)

glue::glue("{nrow(sig_beat)} significant HVGs at FDR < 0.01")
```

```{r}
# Bar plot — top genes by F-statistic
results_beat %>%
  filter(adj.P.Val < 0.01) %>%
  ggplot(aes(x = reorder(hugo, F), y = F)) +
  geom_col(fill = "#4DBBD5") +
  coord_flip() +
  theme_bw() +
  labs(x     = NULL,
       y     = "F-statistic",
       title = "Significant HVGs by Disease Type — BEAT-AML")
```

```{r, fig.width=12, fig.height=9}
long_aml_metab %>%
    filter(cohort == "BEAT", hugo %in% sig_beat$hugo) %>%
    mutate(Disease_Type = fct_collapse(Disease_Type,
                                       therapy_related = c("t-AML", "t-MNs")),
           hugo = factor(hugo, levels = sig_beat$hugo)) %>%
    ggplot(aes(x = Disease_Type, y = log2_tmm_cpm, fill = Disease_Type)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.size = 0.5) +
    facet_wrap(~ hugo, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = NULL, y = "log2(TMM-CPM)")
```

## WGCNA

```{r}
wes_beat <- read_tsv("../../data/processed/wes_wide.tsv",
                     show_col_types = FALSE) %>%
  filter(sample_id %in% meta_beat$sample_id)   # keep only BEAT-AML samples

# Traits matrix — Disease_Type + Gender + WES mutations
traits <- meta_beat %>%
  select(sample_id, Disease_Type, Gender) %>%
  mutate(
    is_sec             = as.integer(Disease_Type %in% "sec"),
    is_therapy_related = as.integer(Disease_Type %in% c("t-AML", "t-MNs")),
    is_female          = as.integer(Gender == "FEMALE")
  ) %>%
  select(sample_id, is_sec, is_therapy_related, is_female) %>%
  inner_join(wes_beat, by = "sample_id") %>%
  mutate(across(where(is.logical), as.integer)) %>%  # TRUE/FALSE -> 1/0
  column_to_rownames("sample_id")

# Expression matrix aligned to traits
mat_wgcna <- mat_beat[rownames(traits), gem$hugo]

# Sanity check
identical(rownames(mat_wgcna), rownames(traits))
glue::glue("Samples: {nrow(mat_wgcna)} | Genes: {ncol(mat_wgcna)} | Traits: {ncol(traits)}")
```


