---
title: "prepare_data"
format: html
editor: source
---

Reads raw files from all three cohorts, aligns samples to WES IDs, and writes clean matrices to `processed/`.

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(readxl)
library(ggVennDiagram)
```

## WES mutations

Long-format somatic mutation table shared across BEAT-AML, TCGA, and NGS-PTL. TCGA sample IDs have their `-03` suffix stripped to match the format used in the expression data.

```{r}
wes <- read_xlsx("../raw/WES_filtered_final.xlsx") |>
  mutate(ID = if_else(cohort == "TCGA", str_remove(ID, "-03"), ID))
```

## BEAT-AML expression

The expression matrix uses RNA-seq IDs (`dbgap_rnaseq_sample`). The sample mapping table (source: [Bottomly et al. 2022](https://github.com/biodev/beataml2.0_data)) converts these to WES IDs (`labId`), keeping only samples flagged for inclusion.

```{r}
expr_beat <- read_tsv("../raw/210126_BEAT-AML_CPM-TMM_log2_with_hugo.tsv.xz",
                      show_col_types = FALSE) |>
  column_to_rownames("symbol") |>
  t() |>
  as_tibble(rownames = "dbgap_rnaseq_sample")

meta_beat <- read_xlsx("../raw/beataml_waves1to4_sample_mapping.xlsx") |>
  filter(rna_include_in_analysis == "yes")

expr_beat <- expr_beat |>
  left_join(select(meta_beat, dbgap_rnaseq_sample, labId),
            by = "dbgap_rnaseq_sample") |>
  relocate(labId, .after = "dbgap_rnaseq_sample") |>
  select(-dbgap_rnaseq_sample)
```

Keep only samples present in both expression and WES.

```{r, fig.width=4, fig.height=3.5}
sample_id_wes <- wes |> filter(cohort == "BEAT-AML") |> pull(ID) |> unique()
ggVennDiagram(x = list(wes = sample_id_wes, rna = expr_beat$labId))
rm(sample_id_wes)
```

```{r}
expr_beat_aml <- expr_beat |>
  filter(labId %in% intersect(unique(wes$ID), labId))
```

## TCGA-AML expression

The raw file has one row per Ensembl ID with a `hugo` column. Rows with missing or duplicated Hugo symbols are removed before pivoting to wide format (samples × genes).

```{r}
expr_tcga <- read_tsv("../raw/200221_TCGA-LAML_CPM-TMM.tsv.xz",
                      show_col_types = FALSE)
```

Inspect duplicated Hugo symbols before dropping them.

```{r}
expr_tcga |>
  group_by(hugo) |>
  filter(n() > 1) |>
  relocate(hugo) |>
  arrange(hugo)
```

```{r}
taxa_tcga <- expr_tcga |> select(hugo, ensembl_gene_id)

expr_tcga <- expr_tcga |>
  filter(!is.na(hugo)) |>
  group_by(hugo) |>
  filter(n() == 1) |>
  ungroup() |>
  select(-ensembl_gene_id) |>
  column_to_rownames("hugo") |>
  t() |>
  as_tibble(rownames = "sample_id")
```

Keep only samples present in both expression and WES.

```{r, fig.width=4, fig.height=3.5}
sample_id_wes <- wes |> filter(cohort == "TCGA") |> pull(ID) |> unique()
ggVennDiagram(x = list(wes = sample_id_wes, rna = expr_tcga$sample_id))
rm(sample_id_wes)
```

```{r}
expr_tcga_aml <- expr_tcga |>
  filter(sample_id %in% intersect(unique(wes$ID), sample_id))
```

## Gene overlap: BEAT-AML vs TCGA-AML

The two cohorts were normalised independently and cannot be merged. Genes exclusive to BEAT-AML are nearly all zero (>99% zero counts) and will be dropped when filtering to Human-GEM metabolic genes.

```{r, fig.width=4, fig.height=3.5}
ggVennDiagram(x = list(beat = colnames(expr_beat_aml)[-1],
                       tcga = colnames(expr_tcga_aml)[-1]))
```

```{r}
inter     <- intersect(colnames(expr_beat_aml)[-1], colnames(expr_tcga_aml)[-1])
only_beat <- setdiff(colnames(expr_beat_aml)[-1], colnames(expr_tcga_aml)[-1])

cat("Shared gene expression fraction — BEAT:", sum(expr_beat_aml[, inter]) / sum(expr_beat_aml[, -1]), "\n")
cat("Shared gene expression fraction — TCGA:", sum(expr_tcga_aml[, inter]) / sum(expr_tcga_aml[, -1]), "\n")
cat("Zero fraction in BEAT-only genes:", sum(expr_beat_aml[, only_beat] == 0) / length(as.matrix(expr_beat_aml[, only_beat])), "\n")

rm(inter, only_beat)
```

## Export expression matrices

```{r}
expr_beat_aml |>
  rename(sample_id = labId) |>
  write_tsv(xzfile("../processed/beat_aml_expression.tsv.xz"))

expr_tcga_aml |>
  write_tsv(xzfile("../processed/tcga_aml_expression.tsv.xz"))
```

## Metadata

Sample-level metadata assembled from the karyotype file and the BEAT-AML mapping table. Rows are sorted to match the expression matrix row order.

```{r, message=FALSE}
karyotype <- read_xlsx("../raw/Karyotype and other variables.xlsx", col_types = "text") |>
  select(-2, -6) |>
  rename(Disease_Type = Disease.type, sample_id = ID)
```

```{r}
beat_aml_metadata <- meta_beat |>
  filter(labId %in% expr_beat_aml$labId) |>
  relocate(labId) |>
  rename(sample_id = labId) |>
  left_join(
    karyotype |>
      filter(cohort == "BEAT-AML") |>
      select(sample_id, Patient_Id, Gender, Disease_Type),
    by = "sample_id"
  ) |>
  arrange(match(sample_id, expr_beat_aml$labId))

tcga_aml_metadata <- karyotype |>
  filter(cohort == "TCGA") |>
  filter(sample_id %in% expr_tcga_aml$sample_id) |>
  arrange(match(sample_id, expr_tcga_aml$sample_id))

write_tsv(beat_aml_metadata,  "../processed/beat_aml_metadata.tsv")
write_tsv(tcga_aml_metadata, "../processed/tcga_aml_metadata.tsv")
```

## Human-GEM: metabolic gene annotation

Human-GEM links genes to metabolic subsystems through GPR (gene–protein–reaction) rules encoded as Boolean expressions with Ensembl IDs. We extract a flat gene → subsystem table used downstream to filter expression matrices to metabolic genes before WGCNA.

```{r}
gem_genes <- read_xlsx("../raw/Human-GEM.xlsx", sheet = "GENES") |>
  select(ensembl_id = NAME, hugo = `SHORT NAME`)

gem_rxns <- read_xlsx("../raw/Human-GEM.xlsx", sheet = "RXNS") |>
  select(reaction_id = ID,
         reaction_name = NAME,
         gene_association = `GENE ASSOCIATION`,
         subsystem = SUBSYSTEM) |>
  filter(!is.na(gene_association))

gem <- gem_rxns |>
  mutate(ensembl_id = str_extract_all(gene_association, "ENSG\\d+")) |>
  unnest(ensembl_id) |>
  left_join(gem_genes, by = "ensembl_id") |>
  filter(!is.na(hugo)) |>
  select(hugo, ensembl_id, reaction_id, reaction_name, subsystem) |>
  distinct()

gem_by_gene <- gem |>
  group_by(hugo, ensembl_id) |>
  summarise(
    n_reactions  = n(),
    n_subsystems = n_distinct(subsystem),
    subsystems   = paste(unique(subsystem), collapse = ";"),
    reactions    = paste(unique(reaction_id), collapse = ";"),
    .groups = "drop"
  )

write_tsv(gem_by_gene, "../processed/gem_gene_subsystems.tsv")
```

Metabolic gene coverage in each cohort:

```{r}
cat("BEAT-AML:", length(intersect(colnames(expr_beat_aml), gem_genes$hugo)), "metabolic genes\n")
cat("TCGA-LAML:", length(intersect(colnames(expr_tcga_aml), gem_genes$hugo)), "metabolic genes\n")
```

## NGS-PTL metabolomics

Log-transformed, DNA-normalised metabolite abundance matrix (sheet: *Log Transformed DNANormImpData*). Samples whose ID starts with `CONTROL` are healthy CD34+ progenitors; all others are AML.

```{r}
metabolome_raw <- read_xlsx("../raw/Metabolome.xlsx",
                            sheet = "Log Transformed DNANormImpData") |>
  rename(metabolite = 1) |>
  filter(metabolite != "Metabolite") |>
  column_to_rownames("metabolite")

ngs_ptl_metabolome <- metabolome_raw |>
  t() |>
  as_tibble(rownames = "sample_id")

ngs_ptl_metadata <- tibble(sample_id = colnames(metabolome_raw)) |>
  mutate(sample_type = if_else(str_starts(sample_id, "CONTROL"), "control", "AML"))

write_tsv(ngs_ptl_metabolome, "../processed/ngs_ptl_metabolome.tsv")
write_tsv(ngs_ptl_metadata,   "../processed/ngs_ptl_metadata.tsv")
```
