---
title: "prepare_data"
format: html
editor: source
---

Reads raw files from all three cohorts, aligns samples to WES IDs, and writes clean matrices to `processed/`.

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(readxl)
library(ggVennDiagram)
```

## WES mutations

Long-format somatic mutation table shared across BEAT-AML, TCGA, and NGS-PTL. TCGA sample IDs have their `-03` suffix stripped to match the format used in the expression data.

```{r}
wes <- read_xlsx("../raw/WES_filtered_final.xlsx") |>
  mutate(ID = if_else(cohort == "TCGA", str_remove(ID, "-03"), ID))
```

## BEAT-AML expression

The expression matrix uses RNA-seq IDs (`dbgap_rnaseq_sample`). The sample mapping table (source: [Bottomly et al. 2022](https://github.com/biodev/beataml2.0_data)) converts these to WES IDs (`labId`), keeping only samples flagged for inclusion.

```{r}
expr_beat <- read_tsv("../raw/210126_BEAT-AML_CPM-TMM_log2_with_hugo.tsv.xz",
                      show_col_types = FALSE) |>
  column_to_rownames("symbol") |>
  t() |>
  as_tibble(rownames = "dbgap_rnaseq_sample")

meta_beat <- read_xlsx("../raw/beataml_waves1to4_sample_mapping.xlsx") |>
  filter(rna_include_in_analysis == "yes")

expr_beat <- expr_beat |>
  left_join(select(meta_beat, dbgap_rnaseq_sample, labId),
            by = "dbgap_rnaseq_sample") |>
  relocate(labId, .after = "dbgap_rnaseq_sample") |>
  select(-dbgap_rnaseq_sample)
```

Keep only samples present in both expression and WES.

```{r, fig.width=4, fig.height=3.5}
sample_id_wes <- wes |> filter(cohort == "BEAT-AML") |> pull(ID) |> unique()
ggVennDiagram(x = list(wes = sample_id_wes, rna = expr_beat$labId))
rm(sample_id_wes)
```

```{r}
expr_beat_aml <- expr_beat |>
  filter(labId %in% intersect(unique(wes$ID), labId))
```

## TCGA-AML expression

The raw file has one row per Ensembl ID with a `hugo` column. Rows with missing or duplicated Hugo symbols are removed before pivoting to wide format (samples × genes).

```{r}
expr_tcga <- read_tsv("../raw/200221_TCGA-LAML_CPM-TMM.tsv.xz",
                      show_col_types = FALSE)
```

Inspect duplicated Hugo symbols before dropping them.

```{r}
expr_tcga |>
  group_by(hugo) |>
  filter(n() > 1) |>
  relocate(hugo) |>
  arrange(hugo)
```

```{r}
taxa_tcga <- expr_tcga |> select(hugo, ensembl_gene_id)

expr_tcga <- expr_tcga |>
  filter(!is.na(hugo)) |>
  group_by(hugo) |>
  filter(n() == 1) |>
  ungroup() |>
  select(-ensembl_gene_id) |>
  column_to_rownames("hugo") |>
  t() |>
  as_tibble(rownames = "sample_id")
```

Keep only samples present in both expression and WES.

```{r, fig.width=4, fig.height=3.5}
sample_id_wes <- wes |> filter(cohort == "TCGA") |> pull(ID) |> unique()
ggVennDiagram(x = list(wes = sample_id_wes, rna = expr_tcga$sample_id))
rm(sample_id_wes)
```

```{r}
expr_tcga_aml <- expr_tcga |>
  filter(sample_id %in% intersect(unique(wes$ID), sample_id))
```

## Gene overlap: BEAT-AML vs TCGA-AML

The two cohorts were normalised independently and cannot be merged. Genes exclusive to BEAT-AML are nearly all zero (>99% zero counts) and will be dropped when filtering to Human-GEM metabolic genes.

```{r, fig.width=4, fig.height=3.5}
ggVennDiagram(x = list(beat = colnames(expr_beat_aml)[-1],
                       tcga = colnames(expr_tcga_aml)[-1]))
```

```{r}
inter     <- intersect(colnames(expr_beat_aml)[-1], colnames(expr_tcga_aml)[-1])
only_beat <- setdiff(colnames(expr_beat_aml)[-1], colnames(expr_tcga_aml)[-1])

cat("Shared gene expression fraction — BEAT:", sum(expr_beat_aml[, inter]) / sum(expr_beat_aml[, -1]), "\n")
cat("Shared gene expression fraction — TCGA:", sum(expr_tcga_aml[, inter]) / sum(expr_tcga_aml[, -1]), "\n")
cat("Zero fraction in BEAT-only genes:", sum(expr_beat_aml[, only_beat] == 0) / length(as.matrix(expr_beat_aml[, only_beat])), "\n")

rm(inter, only_beat)
```

## Export expression matrices

```{r}
expr_beat_aml |>
  rename(sample_id = labId) |>
  write_tsv(xzfile("../processed/beat_aml_expression.tsv.xz"))

expr_tcga_aml |>
  write_tsv(xzfile("../processed/tcga_aml_expression.tsv.xz"))
```

## Metadata

Sample-level metadata assembled from the karyotype file and the BEAT-AML mapping table. Rows are sorted to match the expression matrix row order.

```{r, message=FALSE}
karyotype <- read_xlsx("../raw/Karyotype and other variables.xlsx", col_types = "text") |>
  select(-2, -6) |>
  rename(Disease_Type = Disease.type, sample_id = ID)
```

```{r}
beat_aml_metadata <- meta_beat |>
  filter(labId %in% expr_beat_aml$labId) |>
  relocate(labId) |>
  rename(sample_id = labId) |>
  left_join(
    karyotype |>
      filter(cohort == "BEAT-AML") |>
      select(sample_id, Patient_Id, Gender, Disease_Type),
    by = "sample_id"
  ) |>
  arrange(match(sample_id, expr_beat_aml$labId))

tcga_aml_metadata <- karyotype |>
  filter(cohort == "TCGA") |>
  filter(sample_id %in% expr_tcga_aml$sample_id) |>
  arrange(match(sample_id, expr_tcga_aml$sample_id))

write_tsv(beat_aml_metadata,  "../processed/beat_aml_metadata.tsv")
write_tsv(tcga_aml_metadata, "../processed/tcga_aml_metadata.tsv")
```

## Human-GEM: metabolic gene annotation

Human-GEM links genes to metabolic subsystems through GPR (gene–protein–reaction) rules encoded as Boolean expressions with Ensembl IDs. We extract a flat gene → subsystem table used downstream to filter expression matrices to metabolic genes before WGCNA.

```{r}
gem_genes <- read_xlsx("../raw/Human-GEM.xlsx", sheet = "GENES") |>
  select(ensembl_id = NAME, hugo = `SHORT NAME`)

# Official annotation (gene names, UniProt, compartment)
gem_genes_official <- read_tsv("../raw/Human-GEM-official.tsv",
                                show_col_types = FALSE) |>
  select(ensembl_id  = genes,
         gene_name   = geneNames,
         uniprot     = geneUniProtID,
         compartment = compartments)

gem_rxns <- read_xlsx("../raw/Human-GEM.xlsx", sheet = "RXNS") |>
  select(reaction_id      = ID,
         reaction_name    = NAME,
         gene_association = `GENE ASSOCIATION`,
         subsystem        = SUBSYSTEM) |>
  filter(!is.na(gene_association))

pathway_map <- tribble(
  ~subsystem, ~pathway_class,
  "Glycolysis / Gluconeogenesis",                                      "Carbohydrate metabolism",
  "Starch and sucrose metabolism",                                      "Carbohydrate metabolism",
  "Galactose metabolism",                                               "Carbohydrate metabolism",
  "Fructose and mannose metabolism",                                    "Carbohydrate metabolism",
  "Pentose and glucuronate interconversions",                           "Carbohydrate metabolism",
  "Pyruvate metabolism",                                                "Carbohydrate metabolism",
  "Propanoate metabolism",                                              "Carbohydrate metabolism",
  "Butanoate metabolism",                                               "Carbohydrate metabolism",
  "Pentose phosphate pathway",                                          "Carbohydrate metabolism",
  "Pentose Phosphate Pathway",                                          "Carbohydrate metabolism",
  "Tricarboxylic acid cycle and glyoxylate/dicarboxylate metabolism",   "Carbohydrate metabolism",
  "Glyoxylate and dicarboxylate metabolism",                           "Carbohydrate metabolism",
  "Purine metabolism",                                                  "Nucleotide metabolism",
  "Pyrimidine metabolism",                                              "Nucleotide metabolism",
  "Nucleotide metabolism",                                              "Nucleotide metabolism",
  "Aminoacyl-tRNA biosynthesis",                                        "Nucleotide metabolism",
  "Alanine, aspartate and glutamate metabolism",                        "Amino acid metabolism",
  "Arginine and proline metabolism",                                    "Amino acid metabolism",
  "Histidine metabolism",                                               "Amino acid metabolism",
  "Glycine, serine and threonine metabolism",                           "Amino acid metabolism",
  "Lysine metabolism",                                                  "Amino acid metabolism",
  "Tryptophan metabolism",                                              "Amino acid metabolism",
  "Tyrosine metabolism",                                                "Amino acid metabolism",
  "Valine, leucine, and isoleucine metabolism",                         "Amino acid metabolism",
  "Phenylalanine, tyrosine and tryptophan biosynthesis",                "Amino acid metabolism",
  "Cysteine and methionine metabolism",                                 "Amino acid metabolism",
  "Glutathione metabolism",                                             "Amino acid metabolism",
  "Beta-alanine metabolism",                                            "Amino acid metabolism",
  "Metabolism of other amino acids",                                    "Amino acid metabolism",
  "Phenylalanine metabolism",                                           "Amino acid metabolism",
  "Urea cycle",                                                         "Amino acid metabolism",
  "Sulfur metabolism",                                                  "Amino acid metabolism",
  "C5-branched dibasic acid metabolism",                                "Amino acid metabolism",
  "Serotonin and melatonin biosynthesis",                               "Amino acid metabolism",
  "Fatty acid activation (cytosolic)",                                  "Lipid metabolism",
  "Fatty acid activation (endoplasmic reticular)",                      "Lipid metabolism",
  "Fatty acid biosynthesis (even-chain)",                               "Lipid metabolism",
  "Fatty acid biosynthesis (odd-chain)",                                "Lipid metabolism",
  "Fatty acid biosynthesis (unsaturated)",                              "Lipid metabolism",
  "Fatty acid elongation (even-chain)",                                 "Lipid metabolism",
  "Fatty acid elongation (odd-chain)",                                  "Lipid metabolism",
  "Fatty acid desaturation (even-chain)",                               "Lipid metabolism",
  "Fatty acid desaturation (odd-chain)",                                "Lipid metabolism",
  "Fatty acid biosynthesis",                                            "Lipid metabolism",
  "Fatty acid oxidation",                                               "Lipid metabolism",
  "Fatty acid degradation",                                             "Lipid metabolism",
  "Fatty acid metabolism",                                              "Lipid metabolism",
  "Linoleate metabolism",                                               "Lipid metabolism",
  "Lipoic acid metabolism",                                             "Lipid metabolism",
  "Omega-3 fatty acid metabolism",                                      "Lipid metabolism",
  "Omega-6 fatty acid metabolism",                                      "Lipid metabolism",
  "Arachidonic acid metabolism",                                        "Lipid metabolism",
  "\tArachidonic acid metabolism",                                       "Lipid metabolism",
  "Leukotriene metabolism",                                             "Lipid metabolism",
  "Eicosanoid metabolism",                                              "Lipid metabolism",
  "Acyl-CoA hydrolysis",                                                "Lipid metabolism",
  "Carnitine shuttle (cytosolic)",                                      "Lipid metabolism",
  "Carnitine shuttle (mitochondrial)",                                  "Lipid metabolism",
  "Carnitine shuttle (peroxisomal)",                                    "Lipid metabolism",
  "Carnitine shuttle (endoplasmic reticular)",                          "Lipid metabolism",
  "Acylglycerides metabolism",                                          "Lipid metabolism",
  "Beta oxidation of even-chain fatty acids (peroxisomal)",             "Lipid metabolism",
  "Beta oxidation of odd-chain fatty acids (peroxisomal)",              "Lipid metabolism",
  "Beta oxidation of unsaturated fatty acids (n-9) (peroxisomal)",      "Lipid metabolism",
  "Beta oxidation of phytanic acid (peroxisomal)",                      "Lipid metabolism",
  "Beta oxidation of di-unsaturated fatty acids (n-6) (peroxisomal)",   "Lipid metabolism",
  "Beta oxidation of even-chain fatty acids (mitochondrial)",           "Lipid metabolism",
  "Beta oxidation of odd-chain fatty acids (mitochondrial)",            "Lipid metabolism",
  "Beta oxidation of unsaturated fatty acids (n-7) (mitochondrial)",    "Lipid metabolism",
  "Beta oxidation of unsaturated fatty acids (n-7) (peroxisomal)",      "Lipid metabolism",
  "Beta oxidation of unsaturated fatty acids (n-9) (mitochondrial)",    "Lipid metabolism",
  "Beta oxidation of di-unsaturated fatty acids (n-6) (mitochondrial)", "Lipid metabolism",
  "Beta oxidation of poly-unsaturated fatty acids (mitochondrial)",     "Lipid metabolism",
  "Beta oxidation of branched-chain fatty acids (mitochondrial)",       "Lipid metabolism",
  "Terpenoid backbone biosynthesis",                                    "Lipid metabolism",
  "Steroid metabolism",                                                 "Lipid metabolism",
  "Androgen metabolism",                                                "Lipid metabolism",
  "Cholesterol biosynthesis 1 (Bloch pathway)",                         "Lipid metabolism",
  "Cholesterol biosynthesis 2",                                         "Lipid metabolism",
  "Cholesterol biosynthesis 3 (Kandustch-Russell pathway)",             "Lipid metabolism",
  "Cholesterol metabolism",                                             "Lipid metabolism",
  "Estrogen metabolism",                                                "Lipid metabolism",
  "Formation and hydrolysis of cholesterol esters",                     "Lipid metabolism",
  "Sphingolipid metabolism",                                            "Lipid metabolism",
  "Glycerolipid metabolism",                                            "Lipid metabolism",
  "Glycerophospholipid metabolism",                                     "Lipid metabolism",
  "Glucocorticoid biosynthesis",                                        "Lipid metabolism",
  "Prostaglandin biosynthesis",                                         "Lipid metabolism",
  "Ether lipid metabolism",                                             "Lipid metabolism",
  "Bile acid biosynthesis",                                             "Lipid metabolism",
  "Bile acid recycling",                                                "Lipid metabolism",
  "Triacylglycerol synthesis",                                          "Lipid metabolism",
  "Inositol phosphate metabolism",                                      "Lipid metabolism",
  "Phosphatidylinositol phosphate metabolism",                          "Lipid metabolism",
  "Oxidative phosphorylation",                                          "Energy metabolism",
  "ROS detoxification",                                                 "Energy metabolism",
  "Ubiquinone synthesis",                                               "Energy metabolism",
  "Ubiquinone and other terpenoid-quinone biosynthesis",                "Energy metabolism",
  "O-glycan metabolism",                                                "Glycan metabolism",
  "N-glycan metabolism",                                                "Glycan metabolism",
  "Amino sugar and nucleotide sugar metabolism",                        "Glycan metabolism",
  "Glycosphingolipid biosynthesis-ganglio series",                      "Glycan metabolism",
  "Glycosphingolipid biosynthesis-globo series",                        "Glycan metabolism",
  "Glycosphingolipid biosynthesis-lacto and neolacto series",           "Glycan metabolism",
  "Glycosphingolipid metabolism",                                       "Glycan metabolism",
  "Glycosylphosphatidylinositol (GPI)-anchor biosynthesis",             "Glycan metabolism",
  "Chondroitin / heparan sulfate biosynthesis",                         "Glycan metabolism",
  "Chondroitin sulfate degradation",                                    "Glycan metabolism",
  "Heparan sulfate degradation",                                        "Glycan metabolism",
  "Keratan sulfate biosynthesis",                                       "Glycan metabolism",
  "Keratan sulfate degradation",                                        "Glycan metabolism",
  "Blood group biosynthesis",                                           "Glycan metabolism",
  "Nicotinate and nicotinamide metabolism",                             "Vitamin and cofactor metabolism",
  "Pantothenate and CoA metabolism",                                    "Vitamin and cofactor metabolism",
  "Folate metabolism",                                                  "Vitamin and cofactor metabolism",
  "Biotin metabolism",                                                  "Vitamin and cofactor metabolism",
  "Biopterin metabolism",                                               "Vitamin and cofactor metabolism",
  "Ascorbate and aldarate metabolism",                                  "Vitamin and cofactor metabolism",
  "Porphyrin metabolism",                                               "Vitamin and cofactor metabolism",
  "Retinol metabolism",                                                 "Vitamin and cofactor metabolism",
  "Riboflavin metabolism",                                              "Vitamin and cofactor metabolism",
  "Thiamine metabolism",                                                "Vitamin and cofactor metabolism",
  "Vitamin B12 metabolism",                                             "Vitamin and cofactor metabolism",
  "Vitamin B6 metabolism",                                              "Vitamin and cofactor metabolism",
  "Vitamin D metabolism",                                               "Vitamin and cofactor metabolism",
  "Vitamin E metabolism",                                               "Vitamin and cofactor metabolism",
  "Vitamin C metabolism",                                               "Vitamin and cofactor metabolism",
  "Vitamin A metabolism",                                               "Vitamin and cofactor metabolism",
  "Vitamin B2 metabolism",                                              "Vitamin and cofactor metabolism",
  "Heme degradation",                                                   "Vitamin and cofactor metabolism",
  "Protein assembly",                                                   "Protein metabolism",
  "Protein modification",                                               "Protein metabolism",
  "Protein degradation",                                                "Protein metabolism",
  "Peptide metabolism",                                                 "Protein metabolism",
  "Transport reactions",                                                "Transport",
  "Exchange/demand reactions",                                          "Transport",
  "Xenobiotics metabolism",                                             "Xenobiotics",
  "Drug metabolism",                                                    "Xenobiotics",
  "Toluene degradation",                                                "Xenobiotics",
  "octane oxidation",                                                   "Xenobiotics",
  "Alkaloids biosynthesis",                                             "Xenobiotics",
  "Hippurate metabolism",                                               "Xenobiotics",
  "Insect hormone biosynthesis",                                        "Xenobiotics",
  "Dietary fiber binding",                                              "Xenobiotics",
  "Miscellaneous",                                                      "Other",
  "Isolated",                                                           "Other",
  "Pool reactions",                                                     "Other",
  "Artificial reactions",                                               "Other",
  "non_metabolic",                                                      "non_metabolic"
)

gem <- gem_rxns |>
  mutate(ensembl_id = str_extract_all(gene_association, "ENSG\\d+")) |>
  unnest(ensembl_id) |>
  left_join(gem_genes,          by = "ensembl_id") |>
  left_join(gem_genes_official, by = "ensembl_id") |>
  left_join(pathway_map,        by = "subsystem") |>
  filter(!is.na(hugo)) |>
  select(hugo, ensembl_id, gene_name, uniprot, compartment,
         reaction_id, reaction_name, subsystem, pathway_class) |>
  distinct()

gem_by_gene <- gem |>
  group_by(hugo, ensembl_id, gene_name, uniprot, compartment) |>
  summarise(
    n_reactions   = n_distinct(reaction_id),
    n_subsystems  = n_distinct(subsystem),
    subsystems    = paste(unique(subsystem),    collapse = ";"),
    pathway_class = paste(unique(pathway_class), collapse = ";"),
    reactions     = paste(unique(reaction_id),  collapse = ";"),
    .groups = "drop"
  )

cat("Unique metabolic genes:", nrow(gem_by_gene), "\n")
cat("Unique subsystems:     ", n_distinct(gem$subsystem), "\n")
cat("Gene × reaction pairs: ", nrow(gem), "\n")

write_tsv(gem_by_gene, "../processed/human_gem_gene_subsystem_summary.tsv")
```


## NGS-PTL metabolomics

Log-transformed, DNA-normalised metabolite abundance matrix (sheet: *Log Transformed DNANormImpData*). Samples whose ID starts with `CONTROL` are healthy CD34+ progenitors; all others are AML.

```{r}
metabolome_raw <- read_xlsx("../raw/Metabolome.xlsx",
                            sheet = "Log Transformed DNANormImpData") |>
  rename(metabolite = 1) |>
  filter(metabolite != "Metabolite") |>
  column_to_rownames("metabolite")

ngs_ptl_metabolome <- metabolome_raw |>
  t() |>
  as_tibble(rownames = "sample_id")

ngs_ptl_metadata <- tibble(sample_id = colnames(metabolome_raw)) |>
  mutate(sample_type = if_else(str_starts(sample_id, "CONTROL"), "control", "AML"))

write_tsv(ngs_ptl_metabolome, "../processed/ngs_ptl_metabolome.tsv")
write_tsv(ngs_ptl_metadata,   "../processed/ngs_ptl_metadata.tsv")
```
